---
title: "Cyclistic Bike-Share Case Study"
author: "Vincent Balalian"
output: github_document
---
### Our Business Task: to maximize the number of annual memberships...

...(which Cyclistic's finance analysts have concluded are much more profitable than casual riders), by identifying actionable trends among rider data which the marketing team can use to design new, targeted campaigns.<br> 

The director of marketing believes maximizing the number of annual memberships is key to the company's future success. Using the recommendations of this analysis, the marketing team will design new strategies for converting casual riders into annual members. To do that, we must understand 1) how annual members and casual riders differ, 2) why casual riders would buy a membership, and 3) how digital media could affect our marketing tactics. <br>

Final approval of the recommended marketing program goes to the Cyclistic executive team (which is notoriously detail-oriented). In order to provide them with all the information they need to make their decision, the goal of this analysis is to provide all of the relevant insights in a way that's easy to digest without sacrificing attention-to-detail.

### Data Sources
The "Cyclistic" dataset was provided by Google as part of their Coursera Data Analytics course Capstone project, though the data itself has been made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement). 
<br><br>
Despite the fictional nature of the titular company, the data itself reflects the operations of an actual, high-volume bike-share business, which gives us a substantial amount of comprehensive, real-world data to work with. (Any personal or private information has already been removed.)
<br><br>
The data itself can be found [here](https://divvy-tripdata.s3.amazonaws.com/index.html), where csv files are available for each month/period of data, dating back to 2013. For the purposes of this analysis, we'll be using the data from February 2022 to January 2023.

#### Environment Setup
```{r environment setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(skimr)
library(lubridate)
```
#### Data Import
```{r importing data, message=FALSE, warning=FALSE}
setwd("original data")
df <- list.files() %>% 
  read_csv(id="file_name")
```
#### Data Preview
```{r previewing data}
skim_without_charts(df)
```
A precursory glance tells us: 
<br>1) all 12 months of data were successfully imported into a single dataframe, and 
<br>2) aside from a few variables (start and end stations, and end coordinates), the data is complete.
```{r}
head(df)
```
#### Using this data... 
...we can identify trends among casual riders and annual members by looking at *what* they ride (rideable_type), *when* they ride (started_at, ended_at), and *where* they ride (starting and ending locations). We can also expand on the given "start" and "end" data to determine *how long* they ride, and *how far*.

### Data Processing
The dataset is very long (over 5 million rows), but not too wide. Since the scope of this analysis is relatively simple, processing, analysis, and visualization of the data will be done in **R Studio**, which is uniquely suited to the current business task.

Since there's not much cleaning needed to work with this dataset, let's continue transforming the original dataframe into different forms that will be useful for later analysis and visualization.
<br><br>
First, let's see if we can get an accurate conversion of distance. 
<br><br>
According to usgs.gov, distance between latitudes is fairly constant, but distance between longitudes varies depending on the latitude. Let's take a look at the latitudes of each ride and see if there's much variation.
```{r latitude check, message=FALSE}
ggplot(data=df, mapping=aes(x=start_lat)) +
  geom_histogram()
```
<br>
Since the distribution of latitudes is so visibly narrow, we can assume that our distance conversions will be accurate enough for the purposes of this analysis. Therefore, let's create a new variable, ride_distance, in miles.

#### Transforming Data

The following chunk creates a new dataframe, df2, which also calculates values for ride_time (in minutes) based on the available location data.
```{r initial transformation}
df2 <- df %>% 
  mutate(ride_time=as.numeric(abs(ended_at-started_at))/60,
         lat_diff=round(abs(start_lat-end_lat)*69, digits=6), 
         lng_diff=round(abs(start_lng-end_lng)*52, digits=6),
         ride_distance=round(abs(lat_diff+lng_diff), digits=6),
         day_of_week=wday(df$started_at, label=TRUE)
         ) %>%
  select(-file_name, -ride_id, -end_lat, -end_lng, -lat_diff, 
           -lng_diff)
# reorder for a more favorable view
df2 <- df2[,c(1, 10, 11, 12, 13, 2, 3, 5, 7, 4, 6, 8, 9)]
head(df2)
```
Let's check for outliers among ride_time and ride_distance.
```{r}
df2 %>% 
  select(ride_time, ride_distance) %>% 
  skim()
```
It looks like both variables have some outliers which should be removed for the purposes of this analysis. 
```{r}
probabilities <- c(0.01, 0.50, 0.90, 0.95, 0.99, 0.999, 0.9999)
round(quantile(df2$ride_time, probs=probabilities))
round(quantile(df2$ride_distance, probs=probabilities, na.rm=TRUE)) 
```
While the outliers in ride_distance seem restricted to only a few values, the outliers in ride_time are a bit harder to pinpoint. Looking at the percentiles listed above, it's clear ride_time values start to spike above the 99th percentile. Since the purpose of this analysis is to help target the largest group of potential member converts, when we explore ride times and distances, we'll ignore obvious outliers.

### Exploratory Analysis
First, let's take a look at the distributions of the two primary categorical values: **rideable_type**, and **member_casual**. 
```{r}
count(df, rideable_type, member_casual) %>%
  arrange(desc(n))
ggplot(data=df, mapping=aes(x=fct_infreq(rideable_type), fill=member_casual)) + 
  geom_bar(position="dodge") +
  labs(x="rideable_type")
```
<br>
One thing immediately stands out: while annual members are only *somewhat* more likely to ride classic bikes over electric bikes, a noticeable majority of casual riders use *electric bikes* over classic bikes. Also, 100% of the 178251 docked bike rides were made by casual riders.
<br><br>
Let's see how rides break down by time and distance:
```{r message=FALSE}
df2 %>% 
  filter(ride_time < 45, ride_distance < 7) %>%
  sample_n(1000) %>% 
  ggplot(aes(x=ride_time, y=ride_distance)) +
  geom_point(aes(color=member_casual)) +
  geom_smooth()
```
<br>
Members and casual riders appear to ride for similar times and distances. Let's see if that holds true.
```{r}
time_and_distance_averages <- df2 %>% 
  filter(ride_time < 106, ride_distance < 7) %>%
  group_by(member_casual) %>% 
  summarize(mean_rt=mean(ride_time), mean_rd=mean(ride_distance),)
time_and_distance_averages %>% 
  ggplot(aes(x=member_casual, y=mean_rt, fill=member_casual)) +
  labs(y="average ride time") +
  geom_col()
time_and_distance_averages %>% 
  ggplot(aes(x=member_casual, y=mean_rd, fill=member_casual)) +
  labs(y="average ride distance") +
  geom_col()
```
<br>
Surprisingly, the average casual rider appears to ride for significantly more time than their member counterparts (with little difference between average ride distances).
<br><br>
Let's see how the rides break down by day_of_week.
```{r}
ggplot(df2, aes(x=day_of_week, fill=member_casual)) +
  geom_bar(position="dodge")
```
<br>
Right away, it's clear annual members are far more likely to ride during weekdays, while casual riders ride more often on weekends.
<br><br>
Let's see how rides break down by time of day.
```{r}
df2 %>% 
  mutate(started_at=(hour(started_at) + minute(started_at)/60)) %>% 
  ggplot(aes(x=started_at, color=member_casual)) +
  geom_density(linewidth=2) +
  scale_x_continuous(breaks=seq(from=0, to=24, by=2))
```
<br>
Usage by both members and casual riders tends to ramp up throughout the day, with both dropping off significantly after peaking around 6 pm. Annual members see significant usage spikes corresponding with normal commuting hours, while casual riders' usage is conspicuously low during morning commute hours.
<br><br>
Now, let's break down rides by month:
```{r}
ggplot(df2, aes(x=started_at, color=member_casual)) +
  geom_density(linewidth=2) +
  scale_x_datetime(date_breaks = "month", date_labels = "%b") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x="month")
```
<br>
Among all riders, usage peaks during summer and craters in winter; but the difference is more pronounced among casual riders, with usage heavily concentrated among late spring and summer.
<br><br>
Now let's check out our top (starting) stations overall:
```{r}
top_stations_overall <- df2 %>%
  count(start_station_name, sort=T, name="rides") %>%
  drop_na() %>% 
  head(20)
top_stations_overall
top_stations_overall_count <- df2 %>% 
  filter(start_station_name %in% top_stations_overall$start_station_name) %>% 
  count(start_station_name, member_casual, sort=T, name="rides")
```
```{r}
top_stations_overall_count %>% 
  ggplot(aes(x=start_station_name, y=rides, fill=member_casual)) +
  geom_col() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x="starting station")
```
<br>
The "Streeter Dr & Grand Ave" station, which is the top starting station among all riders, is disproportionately favored by casual riders. Other popular stations also skew strongly in favor of casual riders, including "DuSable Lake Shore Dr & Monroe St," "Millenium Park," "Michigan Ave & Oak St," and "DuSable Lake Shore Dr & North Blvd." 

### Recap

Based on this analysis, we've learned a few things about casual riders and how their tendencies differ from those of current members.

#### On average, Casual Riders:

* **Prefer electric bikes** (Annual members prefer classic bikes.)

* **Ride for longer time periods** (with little difference between ride distances). 

* **Ride more often on weekends** (as opposed to members, who ride more during the week).
<br><br>
The trends above have a "leisurely" aspect to them which suggests casual riders tend to use Cyclistic products recreationally, while annual members are more likely to use them for work or school transportation.

#### Also:

* **Usage by both members and casual riders tends to ramp up throughout the day**, and drop off significantly after **peaking around 6 pm**; but while member usage spikes during both morning and evening commuting hours, **usage by casual riders is conspicuously low in the morning**.
<br>
This could mean several things. Casual riders may be less likely to work full-time jobs. Perhaps a higher share of casual riders are college students. It’s also likely that many casual riders are tourists or, as we’ve seen, “weekend warriors.”

* Among all riders, **usage peaks during summer and craters in winter**; but **the difference is more pronounced among casual riders**, with usage heavily concentrated among summer months.

* **Some stations** (like the "Streeter Dr & Grand Ave" station) **are disproportionately favored by casual riders.**
<br>
(Other popular stations that skew strongly in favor of casual riders include "DuSable Lake Shore Dr & Monroe St," "DuSable Lake Shore Dr & North Blvd," "Millenium Park," and "Michigan Ave & Oak St.”)

### Recommendations

In order to achieve our goal of maximizing the number of annual memberships by converting casual riders, I suggest the marketing department incorporate the following into their plans:

#### 1) Concentrate marketing efforts on the more popular time periods among casual riders...

...such as afternoons and **evenings**, during **weekends**, during the months of **May through September**.

#### 2) Conduct in-person marketing campaigns at select stations...

...which, having higher ratios of casual riders, present an opportunity to efficiently target them where they're most likely to be.

#### 3) Highlight electric bikes, and their range/capacity

We know that, unlike current members, casual riders prefer electric bikes over classic bikes. We also know casual riders ride for substantially longer times. An effective marketing campaign should highlight these two features.
<br><br>
Overall, there are clear differences among trends between casual riders and established members, and clear opportunities to target them using these insights.
<br><br>




